### 2025-01-16 reviewer-response.md

*The cost on line 260 uses $\hat{r}$, which is a mean radius. Now that cost is in
expectation (in addition to the balanced tree hypothesis).*

This is correct; we added justification for our approach to complexity analyses in Sections 1.2, 2.2.2, and 2.2.5.

*Another example, line 326, comes from the minimum with value 2 on line 284. If
the assumption on the LFD is not satisfied, then an extra log factor would
appear in the cost.*

We have, in Theorem 2.1, explicitly stated the assumption that the query is drawn from the same distribution as the other points in the dataset.
This should ensure that, in practical applications with real-world datasets, the LFD around a query is similar to the LFD around its nearest neighbors.
In the theoretical worst case where the LFD around a query is extremely low, repeating tree-search with doubling radii might not encounter any additional points, we may introduce an extra factor (logarithmic in the distance between the query and the k-th nearest neighbor) the in contribution of the tree search to the overall complexity.
However, we can easily detect and sidestep this issue in our implementation by calculating the distance from the query to the center of the root cluster and using that distance, along with the radius and LFD of the root cluster, to determine an initial radius for tree-search.
This approach, once again, brings us to a constant factor for the contribution from tree-search.

### 2025-01-16 sections/4-results.tex
lots more upadtes

Since the time complexity of CAKES algorithms scales with the LFD of the dataset, we examine the LFD of each dataset we used for benchmarks.
Figure~\ref{fig:results:lfd-plots} illustrates the trends in LFD for Fashion-MNIST, Glove-25, Sift, Random, Silva 18S, and RadioML.
In this section, when we discuss trends in LFD, unless otherwise noted, we are referring to the 95$^{th}$ percentile of LFD.
% This is because our algorithms scale exponentially in the LFD, a

@@

Figures~\ref{fig:results:fashion-mnist-scaling},~\ref{fig:results:glove-25-scaling},~and~\ref{fig:results:sift-scaling} show the scaling behavior of CAKES algorithms and existing algorithms on augmented versions of Fashion-MNIST under Euclidean distance,
Glove-25 under cosine distance, and
Sift under Euclidean distance.
Figure~\ref{fig:results:random-scaling} shows the scaling behavior of CAKES algorithms and existing algorithms on a completely randomly generated dataset with the same cardinality and dimensionality as Sift (1 million points in 128 dimensions).
Figures~\ref{fig:results:silva-scaling} and ~\ref{fig:results:radioml-scaling} show the scaling behavior of CAKES on Silva and RadioML respectively.
For these datasets, we took random sub-samples of the full datasets instead of augmenting them to higher cardinalities.
The horizontal axis in each figure shows the cardinality of the dataset augmented with synthetic points (see Section \ref{sec:methods:synthetic-data}).
The left-most point on each line is at the cardinality of the original dataset.
The vertical axis denotes throughput in queries per second.
Both axes are on a logarithmic scale.
In this section, we report results only for $k$-NN search with $k = 10$, but similar plots for $k = 100$ can be found in the Supplement.
For HNSW and ANNOY, we report the recall for each measurement in the plots. For algorithms that exhibit recall greater than $0.9995$, we do not report the recall in the plots, but we do report it in the tables below.
For some algorithms, we were not able to take measurements for each cardinality because the index-building required more RAM than was available (i.e.,\,more then 512GB).
For linear search with CAKES, we only report the throughput for a few of the initial multipliers because the trend is clear.

@@

% Figure~\ref{fig:results:scaling-plots} shows that for Fashion-MNIST, Glove-25, and Sift, as cardinality increases, the CAKES algorithms (Depth-First Sieve in blue, Repeated $\rho$-NN in green, and Breadth-First Sieve in purple) become faster than our Rust implementation of na\"{i}ve linear search (in orange).
% Though we observe this trend, we note that the exact cardinality at which CAKES's algorithms overtake linear search differs by dataset (see Figure~\ref{fig:results:scaling-plots}).

% Which one of the three CAKES algorithms is fastest also differs by dataset.
% For Fashion-MNIST, Depth-First Sieve is consistently fastest, while for Glove-25, the fastest is Repeated $\rho$-NN, and for Sift, Breadth-First Sieve.
% On all three datasets, Depth-First Sieve and Breadth-First Sieve appear to have performance that is \textit{constant} in the cardinality of the dataset.
% With Glove-25, Repeated $\rho$-NN exhibits similar constant scaling.
% We also observe that on these three datasets, for nearly all cardinalities, all three of the CAKES algorithms are faster than FAISS-Flat (in brown), and that at some cardinality, CAKES's algorithms become faster than FAISS-IVF (in pink).

% While recall for CAKES's does \emph{not} degrade with cardinality, recall for HNSW and ANNOY degrades with cardinality.
% In contrast, CAKES has perfect recall on Fashion-MNIST and Sift (where the distance function is a metric), and near-perfect recall on Glove-25 (cosine distance is not a metric).

% In contrast with the results on the ANN Benchmark datasets reported above, as seen in Figure~\ref{fig:results:random-scaling}, we observe performs quite slowly on the Random dataset.
% As with the real datasets, HNSW and ANNOY are the fastest algorithms, and CAKES exhibits perfect recall at all cardinalities.
% HNSW and ANNOY exhibit \textit{much} lower recall on this random dataset than on any of the ANN benchmark datasets,as shown in Tables~\ref{tab:results:qps-and-recall-fmn},~\ref{tab:results:qps-and-recall-glove},~\ref{tab:results:qps-and-recall-sift}, and~\ref{tab:results:qps-and-recall-random}.

For Silva and RadioML, we benchmarked only CAKES's algorithms because HNSW, ANNOY and FAISS support neither the required distance functions nor, in the case of RadioML, complex-valued data.
Due to the massive sizes of these datasets and challenges in generating plausible augmentations, we took random sub-samples ranging up to the entirety of the dataset---rather than augmented versions of the dataset---to examine how performance scales with cardinality.

\begin{figure}
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.99\textwidth]{plots/fashion-mnist_PermutedBall_10_throughput.png}
        \subcaption{Fashion-MNIST for $k=10$.}
        \label{fig:results:fashion-mnist-scaling}
    \end{subfigure}%
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.99\textwidth]{plots/glove-25_PermutedBall_10_throughput.png}
        \subcaption{Glove-25 for $k=10$.}
        \label{fig:results:glove-25-scaling}
    \end{subfigure}%
    \\
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.99\textwidth]{plots/sift_PermutedBall_10_throughput.png}
        \subcaption{Sift for $k=10$.}
        \label{fig:results:sift-scaling}
    \end{subfigure}%
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.99\textwidth]{plots/random_PermutedBall_10_throughput.png}
        \subcaption{A random dataset for $k=10$.}
        \label{fig:results:random-scaling}
    \end{subfigure}%
    \\
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.99\textwidth]{plots/silva-SSU-Ref_PermutedBall_10_throughput.png}
        \subcaption{Silva for $k=10$.}
        \label{fig:results:silva-scaling}
    \end{subfigure}%
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.99\textwidth]{plots/radio-ml_Ball_10_throughput.png}
        \subcaption{RadioML for $k=10$ at SnR = 10dB.}
        \label{fig:results:radioml-scaling}
    \end{subfigure}%
    \\
    \begin{subfigure}[b]{0.94\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{plots/legend.png}
        \label{fig:results:scaling-legend}
    \end{subfigure}%
    \caption{Throughput across six datasets, including a randomly-generated dataset.
    In each plot, the horizontal axis represents increasing cardinality of the dataset, while the vertical axis represents the throughput in queries per second (higher is better).
    For Fashion-MNIST, Glove-25, and Sift, as cardinality increases, the CAKES algorithms become faster than linear search.
    We observe that the cardinality at which CAKES algorithms outperform linear search differs by dataset. 
    % For Fashion-MNIST, CAKES begins outperforming linear search starting at a cardinality near $10^5$, while for Glove-25 and Sift, this happens near $10^6$ and $10^7$ respectively.
    For Fashion-MNIST, Depth-First Sieve is consistently fastest, while for Glove-25, the fastest is Repeated $\rho$-NN, and for Sift, Breadth-First Sieve.
    On all three datasets, Depth-First Sieve and Breadth-First Sieve appear to have performance that is \textit{constant} in the cardinality of the dataset.
    With Glove-25, Repeated $\rho$-NN exhibits similar constant scaling.
    % We also observe that on these three datasets, for nearly all cardinalities, all three of the CAKES algorithms are faster than FAISS-Flat (in brown), and that at some cardinality, CAKES's algorithms become faster than FAISS-IVF (in pink).
    With Silva, as shown in Figure~\ref{fig:results:silva-scaling}, we observe that for all algorithms, throughput initially seems to linearly decrease as cardinality increases, but that it levels off at higher cardinalities.
    Depth-First Sieve is the fastest CAKES algorithm for lower multipliers, but for larger cardinalities, Repeated $\rho$-NN becomes the fastest.
    For RadioML, as shown in Figure~\ref{fig:results:radioml-scaling}, we observe that throughput declines nearly linearly, and that the three CAKES algorithms exhibit virtually indistinguishable performance; in particular, throughput of CAKES's algorithms is identical within three significant figures.
    On all three of these datasets, HNSW and ANNOY are faster than CAKES's algorithms for all cardinalities; however, CAKES exhibits perfect or near-perfect recall on each dataset, while HNSW and ANNOY exhibit much lower recall, as shown in Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove},\ref{tab:results:qps-and-recall-sift}, \ref{tab:results:qps-and-recall-random}.
    }
    \label{fig:results:scaling-plots}
\end{figure}

@@

These four comparisons (referred to as Ball, BalancedBall, PermutedBalancedBall, and PermutedBall, where depth-first-reordering is referred to as Permuted for brevity) for all three search algorithms on the Fashion-MNIST dataset are shown in Figure~\ref{fig:results:distance-counts}. Notably, performance is not only strictly worse for the balanced approaches, but the asymptotic behavior is significantly worse; the instances (such as Depth First Sieve) that seemed to exhibit constant-time behavior in Figure~\ref{fig:results:scaling-plots}, which is effectively PermutedBall, no longer do. We note that on this dataset, depth-first reordering (the Permuted plotlines) seems to have little effect.

>>>>

Since the time complexity of CAKES algorithms scales with the LFD of the dataset, we examine the LFD of each dataset we used for benchmarks.
Figure~\ref{fig:results:lfd-plots} illustrates the trends in LFD for Fashion-MNIST, Glove-25, Sift, Random, Silva 18S, and Radio-ML.
In this section, when we discuss trends in LFD, unless otherwise noted, we are referring to the 95$^{th}$ percentile of LFD.
% This is because our algorithms scale exponentially in the LFD, a

@@

We benchmark the CAKES algorithms, na\"{i}ve linear search, HNSW, ANNOY, and FAISS-IVF on the Fashion-MNIST, Glove-25, Sift, and Random datasets.
We augment these datasets with synthetic points to examine how performance scales with cardinality.
We also benchmark the CAKES algorithms on the Silva and RadioML datasets, and we subsample these datasets instead of augmenting them to examine how performance scales with cardinality.
For each dataset, we use the corresponding distance function (see Table~\ref{tab:datasets:summary}).
For HNSW, ANNOY, and FAISS-IVF, allow for a hyper-parameter search to tune their index for maximum recall.
For CAKES, we build the tree and use our auto-tuning approach (see Section~\ref{sec:methods:auto-tuning}) to select the fastest algorithm for each dataset and cardinality.
We show results for $k=10$ in the main text, but results for $k=100$ can be found in the Supplement.
Figure~\ref{fig:results:scaling-plots} shows the scaling behavior of these algorithms on these datasets while Tables~\ref{tab:results:qps-and-recall-fmn},~\ref{tab:results:qps-and-recall-glove},~\ref{tab:results:qps-and-recall-sift} and~\ref{tab:results:qps-and-recall-random} report the throughput and recall of CAKES's algorithms at each augmented cardinality for Fashion-MNIST, Glove-25, Sift, and Random respectively.

On the Fashion-MNIST, Glove-25, and Sift datasets (in Figures~\ref{fig:results:fashion-mnist-scaling},~\ref{fig:results:glove-25-scaling},~and~\ref{fig:results:sift-scaling} respectively), we observe that as cardinality increases, the Depth-First Sieve algorithm is consistently the fastest CAKES algorithm with a throughput that is constant in the cardinality of the dataset.
The Breadth-First Sieve algorithm is the usually the second fastest, also with a nearly constant throughput across all cardinalities.
The Repeated $\rho$-NN algorithm, however, falls off in throughput as cardinality increases.
All three CAKES algorithms exhibit perfect recall on the Fashion-MNIST and Sift datasets, and near-perfect recall on the Glove-25 dataset.
In contrast, HNSW and ANNOY are faster than CAKES's algorithms for all cardinalities and have near constant throughput as cardinality increases, but their recall degrades quickly as cardinality increases.
FAISS-IVF exhibits linearly decreasing throughput as cardinality increases, which is expected from the algorithm given that we tune the hyper-parameters to maximize.

On the Random dataset, HNSW and ANNOY are still the fastest algorithms but exhibit recall values near 0.
CAKES's algorithms show linearly decreasing throughput as cardinality increases and are also slower than na\"{i}ve linear search.

@@

For Silva and RadioML, we benchmarked only CAKES's algorithms because HNSW, ANNOY and FAISS support neither the required distance functions nor, in the case of RadioML, complex-valued data.
Due to the massive sizes of these datasets and challenges in generating plausible augmentations, we took random sub-samples ranging up to the entirety of the dataset---rather than augmented versions of the dataset---to examine how performance scales with cardinality.
For the Silva dataset (see Figure~\ref{fig:results:silva-scaling}), we observe that throughput initially decreases linearly as cardinality increases, but levels off at higher cardinalities.
Depth-First Sieve is, once again, the fastest CAKES algorithm.
For Radio-ML (see Figure~\ref{fig:results:radioml-scaling}), however, we see that throughput declines nearly linearly with cardinality, and that the three CAKES algorithms are, again, slower that na\"{i}ve linear search.

\begin{figure}
    \captionsetup[subfigure]{aboveskip=-5pt,belowskip=-0.5pt}
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.95\textwidth]{plots/fashion-mnist_PermutedBall_10_throughput.png}
        \subcaption{Fashion-MNIST for $k=10$.}
        \label{fig:results:fashion-mnist-scaling}
    \end{subfigure}%
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.95\textwidth]{plots/glove-25_PermutedBall_10_throughput.png}
        \subcaption{Glove-25 for $k=10$.}
        \label{fig:results:glove-25-scaling}
    \end{subfigure}%
    \\
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.95\textwidth]{plots/sift_PermutedBall_10_throughput.png}
        \subcaption{Sift for $k=10$.}
        \label{fig:results:sift-scaling}
    \end{subfigure}%
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.95\textwidth]{plots/random_PermutedBall_10_throughput.png}
        \subcaption{A random dataset for $k=10$.}
        \label{fig:results:random-scaling}
    \end{subfigure}%
    \\
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.95\textwidth]{plots/silva-SSU-Ref_PermutedBall_10_throughput.png}
        \subcaption{Silva for $k=10$.}
        \label{fig:results:silva-scaling}
    \end{subfigure}%
    \begin{subfigure}[b]{0.5\textwidth}
        \includegraphics[width=0.95\textwidth]{plots/radio-ml_Ball_10_throughput.png}
        \subcaption{RadioML for $k=10$ at SnR = 10dB.}
        \label{fig:results:radioml-scaling}
    \end{subfigure}%
    \\
    \begin{subfigure}[b]{0.94\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{plots/legend.png}
        \label{fig:results:scaling-legend}
    \end{subfigure}%
    \caption{Throughput across six datasets, including a randomly-generated dataset.
    In each plot, the horizontal axis represents increasing cardinality of the dataset, while the vertical axis represents the throughput in queries per second (higher is better).
    For Fashion-MNIST, Glove-25, and Sift, as cardinality increases, the CAKES algorithms become faster than linear search but the cardinality at which this occurs differs by dataset.
    For Fashion-MNIST and Glove-25, Depth-First Sieve is consistently fastest, while for Sift, Repeated $\rho$-NN is the fastest for smaller cardinalities and Depth-First Sieve is the fastest for larger cardinalities.
    With Silva, we observe that for all algorithms, throughput initially seems to linearly decrease as cardinality increases, but that it starts to level off at higher cardinalities.
    Depth-First Sieve is consistently the fastest algorithm on the Silva dataset.
    For Radio-ML and Random, we see that all three CAKES algorithms are slower than na\"{i}ve linear search, and that their throughput decreases linearly with cardinality.
    HNSW and ANNOY are the fastest algorithms on all four datasets we benchmarked them on, but their recall degrades quickly as cardinality increases on all datasets.
    In particular, on the Random dataset, HNSW and ANNOY have near-zero recall.}
    \label{fig:results:scaling-plots}
\end{figure}

@@

These four comparisons (referred to as Ball, BalancedBall, PermutedBall, and, PermutedBalancedBall, where depth-first-reordering is referred to as Permuted for brevity) for all three search algorithms on the Fashion-MNIST dataset are shown in Figure~\ref{fig:results:distance-counts}.
Notably, performance is not only strictly worse for the balanced approaches, but the asymptotic behavior is significantly worse;
the instances (such as Depth-First Sieve) that seemed to exhibit constant-time behavior in Figure~\ref{fig:results:scaling-plots}, which is effectively PermutedBall, no longer do under a balanced clustering.
We note that on this dataset, depth-first reordering (the Permuted plotlines) seems to have little effect.

### 2025-01-16 sections/5-discussion.tex
lots more upadtes

We have presented CAKES, a set of three algorithms for fast $k$-NN search that are generic over a variety of distance functions.
CAKES's algorithms are exact when the distance function in use is a metric (see definition in Section~\ref{sec:methods}).
Even under cosine distance, which is not a metric, CAKES's algorithms exhibit nearly perfect recall.
CAKES's algorithms are designed to be most effective when the data uphold the manifold hypothesis, or in other words, when the data are constrained to a low-dimensional manifold even when embedded in a high-dimensional space.
As a consequence, these algorithms do not perform well on data with random distributions, because of the absence of a manifold structure.
In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset we tested exhibits a manifold structure, as quantified by the percentiles of LFD.
As expected, our results from Figure~\ref{fig:results:scaling-plots} indicate that CAKES's algorithms scale sub-linearly with cardinality on datasets with low LFD, and scale linearly on datasets with high LFD.

@@

As the previous paragraph suggests, the fastest CAKES algorithm varies by dataset.
For example, on Glove-25, Repeated $\rho$-NN exhibits the highest throughput of any CAKES algorithm, despite having the lowest throughput of the three on Sift and Fashion-Mnist.
On Silva, Repeated $\rho$-NN is comparable to the other CAKES algorithms at low multipliers, but becomes the fastest at the higher multipliers.
When we view these results in light of the LFD plots in Figure~\ref{fig:results:lfd-plots}, we realize that Repeated $\rho$-NN appears to be the fastest CAKES algorithm on datasets where a large proportion of the data have a very low LFD, such as Glove-25 and Silva.
This trend is unsurprising, given that Repeated $\rho$-NN relies on a low LFD around the query in order to quickly estimate the correct radius for $k$ hits.
These findings suggest that Repeated $\rho$-NN is the most ``sensitive'' to the manifold structure of the dataset.
Exploring this sensitivity is a potential avenue for future study, as it remains to learn at what LFD Repeated $\rho$-NN transitions away from being the fastest CAKES algorithm, and whether this transition is sharp.

@@

Additionally, we intend to improve the data augmentation process described in Section~\ref{sec:methods:synthetic-data} to better preserve the topological structure of the underlying dataset.
In particular, we expect that favoring augmentation along the first principal components of the local manifold may allow for more realistic data augmentation.

We also plan to investigate hierarchical data compression by representing differences at each level of the binary tree, particularly for string or genomic data, where all differences are discrete.
Such an encoding would allow us to store the data in a compressed format, and to perform search on the compressed data without decompressing it.
This would allow us to perform fast search on datasets that are too large to fit in memory.
An exploration of compression and compressed search is another avenue for future work.

>>>>

We have presented CAKES, a set of three algorithms for fast $k$-NN search that are generic over a variety of distance functions.
CAKES's algorithms are exact when the distance function in use is a metric (see definition in Section~\ref{sec:methods}).
Even under cosine distance, which is not a metric, CAKES's algorithms exhibit nearly perfect recall.
CAKES's algorithms are designed to be most effective when the data uphold the manifold hypothesis, or in other words, when the data are constrained to a low-dimensional manifold even when embedded in a high-dimensional space.
CAKES uses an unbalanced clustering algorithm to build a binary tree that better captures the manifold structure of the data and provides for faster search performance.
As a consequence, these algorithms do not perform well on data with random distributions, because of the absence of a manifold structure.
In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset we tested exhibits a manifold structure, as quantified by the percentiles of LFD.
As expected, our results from Figure~\ref{fig:results:scaling-plots} indicate that CAKES's algorithms scale sub-linearly with cardinality on real-world datasets with low LFD, and scale linearly on synthetic datasets with high LFD.

@@

@@

Additionally, we intend to improve the data augmentation process described in Section~\ref{sec:methods:synthetic-data} to better preserve the topological structure of the underlying dataset.
In particular, we expect that favoring augmentation along the first few principal components of the local manifold may allow for more realistic data augmentation.

We also plan to investigate hierarchical data compression by representing differences at each level of the binary tree, particularly for string or genomic data, where all differences are discrete.
Such an encoding would allow us to store the data in a compressed format, and to perform search on the compressed data without decompressing it in its entirety.
This would allow us to perform fast search on datasets that are otherwise too large to fit in memory.
An exploration of compression and compressed search is another avenue for future work.
